---
title: "Unit 1 BIO144 Homework, Practical, and Quiz"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


## Introduction

On this web page you will find the homework, practical exercises and weekly quiz for Unit 1 of BIO144.

All other information for the course is on the OLAT course webpages, and web pages linked to from there: [OLAT BIO144 Info. Hub](https://lms.uzh.ch/auth/RepositoryEntry/17827987490/CourseNode/112943793799740)

In this set of web pages, in the left margin you will see a navigation pane. You can use this to navigate between the different units of the course, and between the different sections within each unit. At the bottom of this navigation pane you will see "Start Over". If you press this, you will return to the start of the tutorial, and all your progress in all questions will be reset. Feel free to do this if you want to start again from the beginning. Your answers and progress in the homework, practical, and weekly quiz are only for your own practice and learning. They will not be marked or graded.

## Homework

All homework is for your own practice and learning. It will not be marked or graded.

**Homework common to all weeks:**

* Review the material from the lecture(s) this week. Make sure you understand the concepts and ideas presented. Ensure you can follow any mathematical or statistical explanations. Ensure you can do the R that is needed for the week's learning objectives.
* When you find things you don't understand or have trouble doing (e.g., in R), make a note of these. You can then ask a teaching assistant (TA) for help during the practical session, and you can ask on the Discussion Forum on OLAT. You can also ask questions during the lectures.

**Homework specific to this week (ideally before the practical):**

* Read the [Previous knowledge](https://opetchey.github.io/BIO144_Course_Information/1.2-previous-knowledge.html) page of the course information website, and work through the [BIO144 Previous Knowledge self-tests](https://rstudio.mnf.uzh.ch:3939/content/6902b39f-65cc-4952-9a24-17cca0fe3ae2). Make a note of anything you are unsure about.
* If you want to use R on your own laptop, install R and RStudio (see the "Getting R & RStudio" section of the self-tests). If you would rather not, you can use the UZH RStudio Server for everything.

**Code walkthrough:** the code of the reaction time analysis in the first lecture is in the code walkthrough script, linked (with its video walkthrough) at the top of the course book chapter [Introduction](https://opetchey.github.io/BIO144_Course_Book/1.1-intro.html). Its last section ("Practical toolbox") shows code that this practical uses.

**Optional further reading:** The data analysis workflow, in a little more detail, on the [Insights from Data website](https://insightsfromdata.io/insights-workflow.html#insights-workflow). (Not examinable, unless it is also in the course book.)

## Practical

The aims of this first practical are that by the end of it you can:

* use RStudio, both on the UZH RStudio Server (which you will use in the exam) and, if you wish, on your own laptop;
* organise your work in an RStudio project, and install and load add-on packages;
* get a data file into R;
* find and fix errors in R code (everyone makes them, all the time!);
* know what previous knowledge is expected, and which parts of it you need to revise.

Work through the parts in order. Don't worry if you don't finish everything in the practical session: finish at home, and ask on the Discussion Forum if you get stuck.

### Part 1: Get working in RStudio (about 15 minutes)
<!-- LO: LO2.1.1, LO1.1.1 -->

**Everyone, please log in to the UZH RStudio Server now**, even if you plan to work on your own laptop: <https://rstudio.mnf.uzh.ch/>. Log in with your UZH username and password (the same as for your UZH email). You will use RStudio Server in the final exam, so it is important to check now that you can log in, and to get help if you can't.

If you will also work on your own laptop, open RStudio there too.

Check that you can:

1. Find the four panes of RStudio: the script editor (top left; open a new script with File > New File > R Script), the console (bottom left), the environment (top right) and files/plots/packages/help (bottom right).
2. Type `1 + 1` in the console and press Enter.
3. Type `my_number <- 1 + 1` in the script, then run that line (put the cursor on the line and press Ctrl+Enter on Windows/Linux, or Cmd+Enter on a Mac). Look for `my_number` in the environment pane.

```{r LO-u1-panes, echo=FALSE}
# LO: LO2.1.1
question(
  "Match each part of RStudio to what it is for.\n\n1. Script editor\n2. Console\n3. Environment\n\nA. Shows the objects (data, results) that currently exist in R's memory.\nB. Where you write, save and edit your code, so that you have a record of your analysis.\nC. Where R runs commands and prints the results (and error messages).",
  answer("1B, 2C, 3A", correct = TRUE),
  answer("1C, 2B, 3A"),
  answer("1A, 2C, 3B"),
  answer("1B, 2A, 3C"),
  correct = "Correct! Write your code in a script (not directly in the console), so that you can save it, re-run it and share it.",
  incorrect = "Try again. Where did the result of 1 + 1 appear? Where did my_number appear after you ran the line in the script?",
  allow_retry = TRUE
)
```

### Part 2: An RStudio project and add-on packages (about 15 minutes)
<!-- LO: LO2.1.3, LO2.1.10 -->

**Make an RStudio project for BIO144.** An RStudio project is a folder that holds everything for one piece of work (scripts, data, results), and that RStudio remembers. When you open the project, R automatically looks for files in that folder, so reading data files "just works".

In RStudio: File > New Project... > New Directory > New Project. Name it `BIO144` and choose where to put it. RStudio will restart in the new project (you will see "BIO144" at the top right). Do this wherever you will work (on RStudio Server, and on your laptop if you use it). Each time you work on BIO144, open this project first (File > Open Project..., or the menu at the top right).

```{r LO-u1-project-why, echo=FALSE}
# LO: LO2.1.10
question(
  "Why is it a good idea to keep each piece of work in an RStudio project?",
  answer("When the project is open, R looks for files in the project folder, so you can read a data file just by its name.", correct = TRUE),
  answer("Everything for the work (scripts, data, results) is in one place, which makes it easy to find, back up and share.", correct = TRUE),
  answer("R runs faster inside a project."),
  answer("Projects stop you from making errors in your code."),
  correct = "Correct!",
  incorrect = "Think about where R looks for files, and about keeping things organised.",
  allow_retry = TRUE
)
```

**Install the add-on packages we will use.** In the console, run:

```{r eval = FALSE, echo = TRUE}
install.packages("tidyverse")
```

This installs a set of packages, including `readr`, `dplyr`, `tidyr` and `ggplot2`. It can take a few minutes. (On RStudio Server they may already be installed.) Then load them with:

```{r eval = FALSE, echo = TRUE}
library(tidyverse)
```

```{r LO-u1-install-load, echo=FALSE}
# LO: LO2.1.3
question(
  "Which statements about add-on packages are correct?",
  answer("You install a package only once on each computer (e.g. with install.packages(), or the Install button in the Packages pane).", correct = TRUE),
  answer("You load a package with library() every time you start R and want to use it.", correct = TRUE),
  answer("You need to run install.packages() at the start of every script."),
  answer("Once a package is installed, its functions are available without loading it."),
  correct = "Correct! Installing gets the package onto your computer (once); loading makes its functions available in the current R session (each time). Put library() lines at the top of your scripts, but not install.packages().",
  incorrect = "Think about the difference between getting a package onto the computer, and making it available in the current R session.",
  allow_retry = TRUE
)
```

### Part 3: Get the data into R (about 15 minutes)
<!-- LO: LO2.1.5, LO2.1.9 -->

In the lecture you recorded your reaction times. The class data are in a file called `reaction_times_2027.csv`. Download the file into your project folder by running this line in the console (with your BIO144 project open):

```{r eval = FALSE, echo = TRUE}
download.file("https://raw.githubusercontent.com/opetchey/BIO144_Practicals_WAs/refs/heads/main/assets/datasets/reaction_times_2027.csv",
              destfile = "reaction_times_2027.csv")
```

Look in the Files pane: you should see the file in your project folder. (You can also click on it and choose "View File" to see what's inside. But **do not open it in Excel and save it**: Excel can change data in ways that cause problems for R.)

Now open a new script, save it in the project folder (e.g. as `unit1_practice.R`), and write and run:

```{r eval = FALSE, echo = TRUE}
library(readr)
class_RTs <- read_csv("reaction_times_2027.csv")
class_RTs
str(class_RTs)
summary(class_RTs)
```

How many rows (participants) and columns (variables) are there? Does that make sense? Are the reaction times numbers (`<dbl>`) and is sex at birth text (`<chr>`)?

```{r LO-u1-file-not-found, echo=FALSE}
# LO: LO2.1.5, LO2.1.9
question(
  "Suppose read_csv() gives the error: 'reaction_times_2027.csv' does not exist in current working directory. What are likely causes?",
  answer("The file name in the code is not exactly the same as the name of the file (check spelling, capital letters and the .csv ending).", correct = TRUE),
  answer("The file is not in the project folder (or the project is not open), so R is looking in a different folder.", correct = TRUE),
  answer("When downloading through a web browser, an extra ending (e.g. .txt) was added to the file name.", correct = TRUE),
  answer("The readr package is not installed."),
  correct = "Correct! This is the most common problem when starting with R. Check the Files pane: is the file there, and is its name exactly what you typed? (If readr were not loaded, the error would instead say could not find function \"read_csv\".)",
  incorrect = "The message says R could not find the file. Think about the file's name, and where R is looking.",
  allow_retry = TRUE
)
```

```{r LO-u1-variable-types, echo=FALSE}
# LO: LO2.1.5
question(
  "Why is it important to check the variable types (e.g. with str()) straight after reading in data?",
  answer("Because if a column of numbers contains something that is not a number (e.g. '250ms' or a comma as decimal separator), R reads the whole column as text, and calculations on it will fail or be wrong.", correct = TRUE),
  answer("Because R often changes the values when it reads a file."),
  answer("It isn't important; R always gets the types right."),
  correct = "Correct! Checking the number of rows and columns and the variable types catches most import problems before they cause confusing errors later.",
  incorrect = "Think about what happens if a single value in a numeric column contains a letter.",
  allow_retry = TRUE
)
```

### Part 4: Fix the reaction time script (about 40 minutes)
<!-- LO: LO2.1.9, LO1.1.5, LO1.1.6, LO2.1.10 -->

We give you a script that does a version of the analysis from the lecture's live demonstration: it reads in the class data, tidies it, makes some graphs, and does a t-test comparing the reaction times of females and males. **But the script contains about ten deliberate errors.** Your task is to find and fix them, so that the script runs from top to bottom.

Some of the code uses functions from the `dplyr`, `tidyr` and `ggplot2` packages. **You will learn how these work in Unit 2.** For now, read the comments in the script to see what each step does. You do not need to understand how these functions work to fix the errors: they are the kinds of mistakes everyone makes (typos, a missing comma or bracket, and so on).

Download the script into your project folder:

```{r eval = FALSE, echo = TRUE}
download.file("https://raw.githubusercontent.com/opetchey/BIO144_Practicals_WAs/refs/heads/main/assets/scripts/reaction_time_analysis_needs_fixing.r",
              destfile = "reaction_time_analysis_needs_fixing.r")
```

Open it from the Files pane. Then run it **one command at a time from the top** (Ctrl/Cmd+Enter runs the current line, or the whole command if it spans several lines). After each one, look at the console. If there is an error, fix it, and run that line again before moving on.

**How to read an error message.** The message is R telling you what it could not do. Read it carefully, and look for names you recognise. Here is what some common messages usually mean:

| Error message (part of it) | Usual cause | What to check |
|---|---|---|
| `could not find function "xyz"` | Typo in the function name, or the package containing the function is not loaded | Spelling; a `library()` line for the package |
| `object 'xyz' not found` | Typo in the name of an object, or the object was never created (perhaps because an earlier line failed) | Spelling and capital letters; did the line that creates it run without error? |
| `... does not exist in current working directory` | Wrong file name, or the file is not in the project folder | The Files pane |
| `unexpected string constant`, `unexpected symbol`, `unexpected ')'` | Missing or extra comma, bracket or quote | The line, and the one before it |
| `Column 'xyz' is not found` | Typo in a variable name (including capital letters) | `names()` of the data |
| The console shows `+` instead of `>` | R is waiting for you to finish the command: usually a missing closing bracket or quote | Press Esc, then fix the line |

There is more about this in the section *Finding and fixing errors* in Chapter 2 of the course book.

**Watch out for errors that give no error message!** At least one of the errors makes the code run without complaint, but does the wrong thing. Read the comments, and check that each step gives the result it should.

```{r quiz-fixing-errors, echo=FALSE}
# LO: LO2.1.9
question(
  "When you ran the code to read in the data you got an error that included the text:
  could not find function read_cvs. What was the problem?",
  answer("You had not installed the readr package"),
  answer("You had not loaded the readr package with library(readr)"),
  answer("There was a typo in the function name; it should be read_csv, not read_cvs", correct = TRUE),
  answer("You should have used read.csv instead of read_csv"),
  answer("R could not find the data file... maybe it was missing or in the wrong folder"),
  correct = "Correct! The function name was misspelled.",
  incorrect = "The error message indicates that R could not find the function read_cvs. Some of the other problems might have caused errors, but they would have given different error messages.",
  allow_retry = TRUE
)
```

```{r LO-u1-error-match, echo=FALSE}
# LO: LO2.1.9
question(
  "Match each error message to its most likely cause.\n\n1. could not find function \"read_cvs\"\n2. 'reaction_time_2027.csv' does not exist in current working directory\n3. object 'class_RT' not found\n4. Error: unexpected string constant\n\nA. A misspelt name of an object (data or result).\nB. A typo in a function name (or the package is not loaded).\nC. A missing comma between items, e.g. in c().\nD. A wrong file name, or the file is not in the project folder.",
  answer("1B, 2D, 3A, 4C", correct = TRUE),
  answer("1A, 2D, 3B, 4C"),
  answer("1B, 2C, 3A, 4D"),
  answer("1D, 2B, 3A, 4C"),
  correct = "Correct! With a little practice, error messages become very helpful.",
  incorrect = "Use the table above.",
  allow_retry = TRUE
)
```

```{r LO-u1-same-message, echo=FALSE}
# LO: LO2.1.9, LO2.1.3
question(
  "Later in the script you get: could not find function \"pivot_longer\". The function name is spelled correctly. What is the problem?",
  answer("The package that contains the function (tidyr) is not loaded: a library(tidyr) line is missing.", correct = TRUE),
  answer("pivot_longer is not a real function."),
  answer("The data file was not found."),
  correct = "Correct! The same message can have different causes: a typo in the function name (as with read_cvs), or the package not being loaded (as here). You can find which package a function is in with ??pivot_longer, or by searching online.",
  incorrect = "If the spelling is right, why else might R not know about a function?",
  allow_retry = TRUE
)
```

```{r LO-u1-silent-error, echo=FALSE}
# LO: LO2.1.9
question(
  "After the step that removes implausible values, nrow() says there are 0 participants left, and the t-test then fails with 'not enough observations'. But the filter line itself gave no error. What is going on?",
  answer("The filter condition is the wrong way round: it keeps values below 50 *and* above 500, which no value can be. The comment says what it should do: keep values above 50 and below 500.", correct = TRUE),
  answer("All the reaction times in the class were implausible."),
  answer("The t-test function is broken."),
  correct = "Correct! This is the most dangerous kind of error: the code runs, but does the wrong thing. That is why we check the result of each step (here, with nrow()) against what we expect.",
  incorrect = "Compare the code of the filter step with the comment above it.",
  allow_retry = TRUE
)
```

**Using an AI assistant to understand an error.** Choose one of the error messages you met. Paste the message (and the line of code) into an AI assistant and ask it to *explain what the message means*, rather than to fix the code for you. Compare its explanation with your own understanding and with the table above. Then fix the code yourself.

```{r LO-u1-ai-use, echo=FALSE}
# LO: LO1.1.6
question(
  "Which of these are appropriate uses of an AI assistant while learning data analysis in BIO144?",
  answer("Asking it to explain what an error message means, then fixing the code yourself.", correct = TRUE),
  answer("Asking it to explain a statistical concept in a different way, and checking its explanation against the course book.", correct = TRUE),
  answer("Pasting in the whole practical and copying its answers without trying yourself first."),
  answer("Using it during the final exam."),
  correct = "Correct! Use AI to help you understand, not to do the learning for you. AI assistants can be wrong, so check what they say. And you will not have one in the exam (Copilot is switched off on RStudio Server in exam mode), so you need to be able to read errors and write code yourself.",
  incorrect = "Think about what helps you learn, and about the exam conditions.",
  allow_retry = TRUE
)
```

**Finally, check the script is reproducible.** When you have fixed all the errors: Session > Restart R, then run the whole script from the top (Code > Run Region > Run All). If it runs without any errors, from a fresh start, then it will give the same results every time, for you and for anyone you share it with.

```{r LO-u1-restart, echo=FALSE}
# LO: LO1.1.5, LO2.1.10
question(
  "Why restart R and run the whole script from the top at the end?",
  answer("To check that the script works on its own, and does not depend on objects you created earlier by running lines out of order or typing in the console.", correct = TRUE),
  answer("Because R becomes slower the longer it runs."),
  answer("To delete the data file."),
  correct = "Correct! A script that runs from top to bottom in a fresh R session is a complete, repeatable record of your analysis.",
  incorrect = "Think about what could be in R's memory that is not created by the script itself.",
  allow_retry = TRUE
)
```

If you got stuck, here is the corrected script. Please only look at it after trying to fix the errors yourself: [reaction_time_analysis_working.r](https://raw.githubusercontent.com/opetchey/BIO144_Practicals_WAs/refs/heads/main/assets/scripts/reaction_time_analysis_working.r)

*If you finish early:* change the code that makes the box plot, so that the axis labels are more informative (hint: add `+ labs(x = "Sex at birth", y = "Mean reaction time (ms)")`). Or look at the t-test output: what is the estimated difference between the groups, and its 95% confidence interval? (We will look at reporting results in later units.)

### Part 5: Previous knowledge check-in (about 10 minutes, or at home)
<!-- LO: LO1.1.2 -->

BIO144 builds on things you learned in previous courses (in particular MAT183 and BIO134): for example descriptive statistics, probability distributions, hypothesis testing and t-tests, and basic R. The full list is on the [Previous knowledge](https://opetchey.github.io/BIO144_Course_Information/1.2-previous-knowledge.html) page.

* If you have not done so yet, work through the [BIO144 Previous Knowledge self-tests](https://rstudio.mnf.uzh.ch:3939/content/6902b39f-65cc-4952-9a24-17cca0fe3ae2).
* Write down the two or three topics you feel least sure about. Ask a TA about one of them now, and plan when you will revise the others (before Unit 3 is a good target).

```{r LO-u1-prior-knowledge, echo=FALSE}
# LO: LO1.1.2
question(
  "Where can you find out what previous knowledge BIO144 expects, and check your own?",
  answer("On the Previous knowledge page of the course information website, and in the BIO144 Previous Knowledge self-tests.", correct = TRUE),
  answer("Only in the final exam."),
  answer("There is no expected previous knowledge."),
  correct = "Correct! None of the self-tests are graded: they are there to help you find what to revise.",
  incorrect = "Look at the links above.",
  allow_retry = TRUE
)
```

## Weekly Quiz

In the final exam you will experience questions that ask for a numeric answer. Then you will need to give the answer with the specified precision. For example, if the question asks for pi rounded to 2 digits, you should answer 3.14. If you answer 3.1 or 3.142 you could be marked wrong. Here is what that might look like in a question. Try giving a wrong answer to see what happens!

```{r quiz2, echo=FALSE}
# LO: LO1.1.1
question_numeric(
  "What is pi rounded to 2 digits?",
  answer(3, message = "Too few digits"),
  answer(3.1, message = "Too few digits"),
  answer(3.142, message = "Too many digits"),
  answer(3.14, correct = TRUE),
  allow_retry = TRUE,
  min = 3,
  max = 4,
  step = 0.01
)
```

In the final exam you will sometimes get multiple choice questions where more than one answer is correct. Here is an example question. Try it out!


```{r quiz3, echo=FALSE}
# LO: LO1.1.1
question_checkbox(
  "Select all the toppings that belong on a Margherita Pizza:",
  answer("tomato", correct = TRUE),
  answer("mozzarella", correct = TRUE),
  answer("basil", correct = TRUE),
  answer("extra virgin olive oil", correct = TRUE),
  answer("pepperoni", message = "Pepperoni is a great topping! ... just not on a Margherita Pizza"),
  answer("onions", message = "Onion is a questionable topping, and definitely not on a Margherita Pizza"),
  answer("bacon", message = "Never put bacon on a pizza, please!"),
  answer("spinach", message = "Spinach is a great topping! ... just not on a Margherita Pizza"),
  random_answer_order = TRUE,
  allow_retry = TRUE
)
```

(Obviously pizza toppings are not part of the BIO144 course content. This question is just to illustrate the question type.)

***Please note that if you selected only two of the correct toppings you get the message "Incorrect. Be sure to select every correct answer." This is because all four correct toppings need to be selected to get the question right. It does not necessarily mean that the options you selected are wrong (though they might be). It means that you might have missed some correct options, and or you might have selected some incorrect options.***

In the final exam there may be more complex types of questions. Here is an example of one such question type. Try it out!

```{r weekly-quiz-1, echo=FALSE}
# LO: LO1.1.3
question(
    "The workflow\n\nIn the workflow we use for data analysis (see the live demonstration in Chapter 1 of the course book), which order should the following come in?\n\nA. Visualisation.\nB. Statistical test.\nC. Question.\nD. Data wrangling (tidying and cleaning).\nE. Planned presentation and analysis.\nF. Data collection.\nG. Reporting and communicating.\n\n(Choose the best ordering.)",
    answer("C, F, D, A, B, E, G"),
    answer("C, E, F, D, A, B, G", correct = TRUE),
    answer("F, C, D, A, B, E, G"),
    answer("C, E, F, D, B, A, G"),
    correct = "Correct! Notice that the presentation and analysis are planned before any data are collected.",
    incorrect = "Don't start anything without a question. Plan how you will present and analyse the data before you collect them. Don't do any statistics before visualising your data.",
    allow_retry = TRUE
  )
```

And now for some more questions about this weeks content. These, like the questions in all the following Weekly Quizzes are just for you to practice with. They do not count towards your final grade. They should help you know if you're following and understanding the material.

```{r weekly-quiz-2, echo=FALSE}
# LO: LO2.1.4
question(
    "Getting help\n\nFor each R function there is a help file. You can get these in R by typing in the Console a question mark (?) followed immediately by the function name. For example, to see the help file for the function t.test, type ?t.test.\n\nGo to the help file for t.test. What is the name of the third argument you can give it?",
    answer("alternative", correct = TRUE),
    answer("method"),
    answer("confidence"),
    answer("var.equal"),
    correct = "Correct!",
    incorrect = "Try again... look in the section of the help file named 'Arguments', and find the third entry.",
    allow_retry = TRUE
  )
```

Some questions in the weekly quizzes, and during the course might appear a bit subjective. They are designed to make you think about the material, and to reflect on your own experience. There are no right or wrong answers to these questions. (There are no questions like this in the final exam, of course!) Here is an example:

```{r weekly-quiz-3, echo=FALSE}
# LO: LO2.1.9
question(
    "Most common problem in R\n\nWhen you were correcting the script, you came across various things that needed fixing. Based on this experience, and your intuition: what is the most common cause of error messages we get from R?",
    answer("Variables changing name by themselves"),
    answer("A bug in R"),
    answer("Small typographical mistakes we make", correct = TRUE),
    answer("R making mistakes"),
    correct = "Correct!",
    incorrect = "Try again, and recall the types of errors you encountered while fixing the script.",
    allow_retry = TRUE
  )
```

Sometimes a weekly quiz question might even need you to look up something new, or something from a previous course. This is to encourage you to develop good habits of looking things up, and learning independently. Here might be an example:

```{r weekly-quiz-5, echo=FALSE}
# LO: LO1.1.2
question(
    "When to use a t-test\n\nWhich of the following situations is most appropriate for using a t-test?",
    answer("Comparing the means of more than two groups"),
    answer("Comparing the means of two groups", correct = TRUE),
    answer("Testing for a relationship between two continuous variables"),
    answer("Measuring the slope of a regression line"),
    correct = "Correct!",
    incorrect = "Try again. For more than two groups, an ANOVA is more appropriate. For relationships between continuous variables, correlation or regression is used. You will learn about these types of analyses later in the course.",
    allow_retry = TRUE
  )
```

Another example:

```{r weekly-quiz-6, echo=FALSE}
# LO: LO2.1.1
question(
    "Sometimes when we run a line of code we get no response from R. What does this usually mean?",
    answer("It depends on whether we see a flashing cursor or not", correct = TRUE),
    answer("If we see a plus sign (+) prompt, it means R is waiting for us to finish the command. We might have left a bracket or quote unclosed.", correct = TRUE),
    answer("If we see the usual > prompt, it means the command has run successfully but there is no output to show. No news is good news", correct = TRUE),
    answer("We asked R to do something that is taking a long time, so we need to wait!", correct = TRUE),
    correct = "Correct! All of these are possible explanations for getting no response from R. We have to look carefully in the Console to see which is the case.",
    allow_retry = TRUE
  )
```

Final question for this week:

```{r weekly-quiz-7, echo=FALSE}
# LO: LO2.1.9, LO1.1.6
question(
  "We will often get error messages from R. What is the best way to deal with these?",
  answer("Panic!", message = "Panic is not helpful. Take a deep breath and read the error message carefully."),
  answer("Copy and paste the error message into an AI assistant and use whatever fix it suggests", message = "An AI assistant can sometimes help explain an error, but if you skip reading the message yourself you don't learn to fix errors, and you won't have an AI assistant in the exam. Use it to understand the error, not to fix it for you."),
  answer("Read the error message carefully, and try to understand what it is telling you. Then check your code for mistakes.", correct = TRUE),
  answer("Ask a teaching assistant (TA) for help immediately", message = "TAs are very happy to help, but first try reading the error message and checking your code. Ask for help if you are still stuck."),
  correct = "Correct! Dealing with error messages is a key skill in programming. Understanding what the error message is telling you, and checking your code carefully (especially for typos), is the best first step. If you are still stuck, ask a TA or on the Discussion Forum. See also the section \"Finding and fixing errors\" in Chapter 2 of the course book.",
  allow_retry = TRUE
)
```

And two questions about the ideas in Chapter 1 of the course book:

```{r LO-four-principles, echo=FALSE}
# LO: LO1.1.4
learnr::question(
  "Match each practice to the principle of high-quality data analysis (Popovic et al. 2024) that it follows.\n\n1. Deciding the question and the planned analysis before collecting the data.\n2. Using a Poisson GLM for counts, and a mixed model when the same individuals are measured repeatedly.\n3. Reporting the size of a difference with its confidence interval, and discussing whether it matters biologically.\n4. Sharing the data and the R script with the paper.\n\nA. Report methods and findings so the research is reproducible.\nB. A focused question, with planned sampling and analysis.\nC. A model that accounts for the distribution and dependence of the data.\nD. Emphasise effect sizes and biological relevance.",
  answer("1B, 2C, 3D, 4A", correct = TRUE),
  answer("1B, 2D, 3C, 4A"),
  answer("1A, 2C, 3D, 4B"),
  answer("1C, 2B, 3D, 4A"),
  correct = "Correct! You will meet all four principles again and again during the course.",
  incorrect = "Look again at the four principles in Chapter 1 of the course book.",
  allow_retry = TRUE
)
```

```{r LO-notation, echo=FALSE}
# LO: LO1.1.7
learnr::question(
  "Which statements about the course notation are correct?",
  answer("s is the standard deviation of a sample; σ (sigma) is the standard deviation of the population.", correct = TRUE),
  answer("A hat shows an estimate: β̂₁ (beta-1-hat) is the estimated slope, and ŷᵢ (y-hat-i) is the fitted value for observation i.", correct = TRUE),
  answer("x̄ (x-bar) is the mean of a sample.", correct = TRUE),
  answer("The residual eᵢ is the fitted value minus the mean of y."),
  correct = "Correct! (The residual eᵢ is the observed value minus the fitted value: yᵢ − ŷᵢ.) The full list is in the section 'Notation and some definitions' in Chapter 1 of the course book.",
  incorrect = "Check the section 'Notation and some definitions' in Chapter 1 of the course book.",
  allow_retry = TRUE
)
```

```{r LO-u1-where-help, echo=FALSE}
# LO: LO1.1.1
question(
  "You are stuck on an R problem at home, two days before the next practical. What is a good way to get help?",
  answer("Post on the BIO144 Discussion Forum on OLAT, including your code, the full error message, and what you have already tried.", correct = TRUE),
  answer("Bring it to the next practical and ask a TA.", correct = TRUE),
  answer("Give up on that part of the course."),
  correct = "Correct! The FAQ page of the course information website explains how to ask a question so that we can help quickly. Posts can be anonymous, and others probably have the same question.",
  incorrect = "Look at the 'Getting help' section of the General information page of the course information website.",
  allow_retry = TRUE
)
```

## Solutions
<!-- 2027 addition: link to the example solution script -->

**Please read this before you open the solutions.**

* You will learn much more if you solve the practical yourself first, ideally together with other students, and ask a teaching assistant when you are stuck. Use the solution script afterwards, to check and compare your own work.
* The solution script shows **one** way to solve the practical. It is not the only way: different code that gives the same answers is just as good.

The example solution script for this practical is here: [unit01_practical_solution.R](https://raw.githubusercontent.com/opetchey/BIO144_Practicals_WAs/refs/heads/main/solutions/unit01_practical_solution.R)

To download it straight into your RStudio project folder, run this line in the Console:

```
download.file("https://raw.githubusercontent.com/opetchey/BIO144_Practicals_WAs/refs/heads/main/solutions/unit01_practical_solution.R", destfile = "unit01_practical_solution.R")
```

Then open it in RStudio and work through it line by line, comparing it with your own script. The comments explain each step and give the answers to the questions in the practical.
