---
title: "Unit 11 BIO144 Course Content"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction

On this web page you will find the homework, practical exercises and weekly quiz for Unit 11 of BIO144.

All other information for the course is on the OLAT course webpages, and web pages linked to from there: [OLAT BIO144 Info. Hub](https://lms.uzh.ch/auth/RepositoryEntry/17827987490/CourseNode/112943793799740)


## Homework

All homework is for your own practice and learning. It will not be marked or graded.

### Homework common to all weeks

* Review the material from the lecture(s) this week. Make sure you understand the concepts and ideas presented. Ensure you can follow any mathematical or statistical explanations. Ensure you can do the R that is needed for the week's learning objectives.

* When you find things you don't understand or have trouble doing (e.g., in R), make a note of these. You can then ask a teaching assistant (TA) for help during the practical session, and you can ask on the Discussion Forum on OLAT. You can also ask questions during the lectures.

### Homework specific to this week

There is a very nice and (relatively simple) walkthrough by [Dr Bodo Winter](https://bodowinter.com/index.html), a lecturer in cognitive linguistics at the University of Birmingham UK, of concepts underlying linear mixed effects models. It provides a (slightly) different perspective from those in Prof. Petchey's lecture. It also very clearly walks through an example in R. [Here is the tutorial.](https://web.archive.org/web/20170328211836/http://www.bodowinter.com/tutorial/bw_LME_tutorial2.pdf)

For a somewhat more advanced treatment of LMMs and some examples of implementing them in R, please have a look at Chapter 10 of The New Statistics with R by Hector.

### The analytic world

Here are a series of videos recorded some time ago by Prof. Petchey, showing a small part of the world of analytical methods. It is not really important that you watch all this now, but you might find it interesting. Perhaps skip to the end and watch the last one first, so you see the whole world and then decide what parts you want to explore further. (And during the final lecture of the course, you heard about even more of this world.)

<iframe width="560" height="315" src="https://www.youtube.com/embed/t2XUNhStJ1M?si=KeHDbL4gaMGfH5Br" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" referrerpolicy="strict-origin-when-cross-origin" allowfullscreen></iframe>

<iframe width="560" height="315" src="https://www.youtube.com/embed/pCa9PctaEi0?si=VBIIA0qLHqqsatPm" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" referrerpolicy="strict-origin-when-cross-origin" allowfullscreen></iframe>

<iframe width="560" height="315" src="https://www.youtube.com/embed/wztQAz-12iY?si=JRNHtv3b_KyeClT2" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" referrerpolicy="strict-origin-when-cross-origin" allowfullscreen></iframe>

<iframe width="560" height="315" src="https://www.youtube.com/embed/T4cGt8OZRgI?si=CTL5WflivzJgBH_q" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" referrerpolicy="strict-origin-when-cross-origin" allowfullscreen></iframe>

<iframe width="560" height="315" src="https://www.youtube.com/embed/HSUhJjRQDRo?si=jyF7XPGWLvNT05Jc" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" referrerpolicy="strict-origin-when-cross-origin" allowfullscreen></iframe>



## Practical part 1
<!-- LO: LO11.1.5, LO11.1.7, LO11.1.8 -->

In the lecture and course book there were various examples and walkthroughs of linear mixed effects models (LMMs). In this practical you will experience a core feature of LMMs: the partial pooling of information across groups. This is one of the key features that distinguishes LMMs from standard linear models (LMs). It is important to understand this feature well, as it is central to the correct use and interpretation of LMMs.

Partial pooling across groups means that the estimates for each group are "shrunk" towards the overall mean. The amount of shrinkage depends on the amount of data available for each group and the variability within and between groups.

Let us break this down into steps.

First, simulate some data with a grouping structure. For example, you could simulate test scores for students in different classrooms, where each classroom has a different average score.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
## set seed for reproducibility
set.seed(123)
## number of classes
n_classes <- 10
## class effects
class_average_score <- runif(n_classes, min = 60, max = 90)
## number of students per class
n_students_per_class <- sample(2:25, n_classes, replace = TRUE)
## class features data frame
class_features <- tibble(
  class = paste0("class_", 1:n_classes),
  n_students = n_students_per_class,
  class_effect = class_average_score
)
## simulate student scores
student_scores <- class_features |> 
  rowwise() |> 
  mutate(
    scores = list(rnorm(n_students, mean = class_effect, sd = 15))
  ) |> 
  unnest(cols = c(scores)) |> 
  select(class, score = scores) |> 
  mutate(student_id = paste0("student_", row_number()))
```

Here is the data we have simulated:

```{r}
student_scores
```

Check the number of students per class:

```{r}
student_scores |>
  group_by(class) |>
  summarise(n_students = n())
```

Now over to you. Make a script, copy and paste the code above so you have the dataset.

Then make a standard linear model with the intention to predict student scores based on class.

```{r}
## standard linear model
lm_model <- lm(score ~ class, data = student_scores)
```


```{r quiz-lmm-lm-what-estimated, echo=FALSE}
# LO: LO11.1.5
learnr::question(
  "In the linear model `lm(score ~ class)`, what does the model estimate for each class?",
  
  learnr::answer(
    "A separate mean score for each class, estimated independently of the other classes.",
    correct = TRUE
  ),
  learnr::answer("A single mean score shared by all classes."),
  learnr::answer("A random deviation around the overall mean."),
  learnr::answer("A slope describing how score changes with class number."),
  
  correct = "Correct! A linear model estimates a separate mean for each class, without sharing information across classes.",
  incorrect = "Think about what is estimated in a linear model... and they are estimated.",
  allow_retry = TRUE
)
```

Make a dataset that contains the means of each class from the linear model.

```{r}
new_data <- data.frame(class = unique(student_scores$class))
new_data$predicted_lm <- predict(lm_model, newdata = new_data)
new_data
```

And now just calculate the mean score in each class with dplyr:

```{r}
class_means <- student_scores |>
  group_by(class) |>
  summarise(mean_score = mean(score))
class_means
```

You should see that the predicted means from the linear model and the calculated means are identical. The linear model is simply estimating the mean score for each class independently.

This is exactly what we expect from a standard linear model. The standard linear model treats the class as a *fixed effect*, estimating a separate mean for each class without sharing information across classes.

Now, let's fit a linear mixed effects model (LMM) to the same data, treating class as a random effect. This means that the model will estimate an overall mean score and allow each class to deviate from this mean.

```{r message=FALSE, warning=FALSE}
library(lme4)
## linear mixed effects model
lmm_model <- lmer(score ~ 1 + (1 | class), data = student_scores)
```

Now, make predictions from the LMM for each class

```{r}
new_data <- new_data |> 
  mutate(predicted_lmm = predict(lmm_model, newdata = new_data, re.form = NULL))
new_data
```



```{r quiz-lmm-lm-vs-lmm-predictions, echo=FALSE}
# LO: LO11.1.7
learnr::question(
  "When you compare the class-level predictions from the standard linear model (LM) and the linear mixed-effects model (LMM), what pattern do you typically observe?",
  
  learnr::answer(
    "The predictions from the LMM are identical to those from the LM, because both models use the same data."
  ),
  learnr::answer(
    "The lower LMM predictions tend to be larger than the corresponding LM predictions, while the higher LMM predictions tend to be smaller.",
    correct = TRUE
  ),
  learnr::answer(
    "The LMM predictions are always lower than the LM predictions for all classes."
  ),
  learnr::answer(
    "The LMM predictions are more extreme than the LM predictions for all classes."
  ),
  
  correct = "Correct! This pattern reflects partial pooling: extreme LM estimates are shrunk toward the overall mean in the LMM.",
  incorrect = "Partial pooling tends to cause shrinkage... think about what that might mean about the distribution of the predictions.",
  allow_retry = TRUE
)
```

Partial pooling means that the estimates for each class are "shrunk" towards the overall mean. The amount of shrinkage depends on the amount of data available for each class and the variability within and between classes.

Let's visualise this effect by plotting the predictions from both models against the calculated class means.

```{r}
library(ggplot2)
ggplot(new_data, aes(x = class)) +
  geom_point(aes(y = predicted_lm, color = "LM Predictions"), size = 3, alpha = 0.5) +
  geom_point(aes(y = predicted_lmm, color = "LMM Predictions"), size = 3, alpha = 0.5) +
  geom_hline(yintercept = mean(student_scores$score), linetype = "dashed", color = "black") +
  labs(y = "Score", color = "Legend") +
  theme_minimal() +
  coord_flip()
```

```{r quiz-lmm-graph-interpretation, echo=FALSE}
# LO: LO11.1.7
learnr::question(
  "Looking at the graph showing class-level predictions from the linear model (LM) and the linear mixed-effects model (LMM), together with a horizontal line indicating the overall mean, what do you observe?",
  
  learnr::answer(
    "The LM and LMM predictions are identical, and both lie exactly on the overall mean line."
  ),
  learnr::answer(
    "The LMM predictions are more spread out than the LM predictions, moving further away from the overall mean."
  ),
  learnr::answer(
    "The LM predictions are all closer to the overall mean than the LMM predictions."
  ),
  learnr::answer(
    "The LMM predictions are closer to the overall mean than the LM predictions, with extreme class estimates pulled toward the mean.",
    correct = TRUE
  ),
  
  correct = "Correct! This is the key visual signature of partial pooling: LMM predictions shrink extreme class estimates toward the overall mean.",
  incorrect = "Focus on how the LMM predictions relate to both the LM predictions and the overall mean line.",
  allow_retry = TRUE
)
```


Look at the number of students in each class (you calculated this above). Which classes were pulled most strongly towards the overall mean? How much a class is shrunk depends on how many students it has and on how variable scores are within classes compared with among classes. It does not depend on how many classes there are.

To see this clearly, here is the same simulation with 1'000 classes. With so many classes, the overall pattern is easy to see.

```{r echo = FALSE, eval = TRUE}
## set seed for reproducibility
set.seed(123)
## number of classes
n_classes <- 1000
## class effects
class_average_score <- runif(n_classes, min = 60, max = 90)
## number of students per class
n_students_per_class <- sample(2:25, n_classes, replace = TRUE)
## class features data frame
class_features <- tibble(
  class = paste0("class_", 1:n_classes),
  n_students = n_students_per_class,
  class_effect = class_average_score
)
## simulate student scores
student_scores <- class_features |> 
  rowwise() |> 
  mutate(
    scores = list(rnorm(n_students, mean = class_effect, sd = 15))
  ) |> 
  unnest(cols = c(scores)) |> 
  select(class, score = scores) |> 
  mutate(student_id = paste0("student_", row_number()))
## standard linear model
lm_model <- lm(score ~ class, data = student_scores)
## linear mixed effects model
lmm_model <- lmer(score ~ 1 + (1 | class), data = student_scores)
## make new data for predictions
new_data <- data.frame(class = unique(student_scores$class))
new_data <- new_data |>
  mutate(
    predicted_lm = predict(lm_model, newdata = new_data),
    predicted_lmm = predict(lmm_model, newdata = new_data, re.form = NULL)
  ) |>
  left_join(select(class_features, class, n_students), by = "class")
```


```{r echo = FALSE, eval = TRUE}
## plot predictions
ggplot(new_data, aes(x = class)) +
  geom_point(aes(y = predicted_lm, color = "LM Predictions"), size = 1, alpha = 0.5) +
  geom_point(aes(y = predicted_lmm, color = "LMM Predictions"), size = 1, alpha = 0.5) +
  geom_hline(yintercept = mean(student_scores$score), linetype = "dashed", color = "black") +
  labs(y = "Score", color = "Legend") +
  coord_flip() +
  ## no tick labels on x axis
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
```

This is art that tells a story about data and models.

And here are the distributions of the predictions from the two models:

```{r echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE}
## plot distributions of predictions
ggplot(new_data) +
  geom_density(aes(x = predicted_lm, color = "LM Predictions"), size = 1) +
  geom_density(aes(x = predicted_lmm, color = "LMM Predictions"), size = 1) +
  labs(x = "Predicted Score", color = "Legend") +
  theme_minimal()
```

And here is how much each class's estimate is shrunk towards the overall mean, against the number of students in that class. A shrinkage of 0 means the LMM estimate equals the class mean (no pooling); a shrinkage of 1 would mean it equals the overall mean (complete pooling). (Classes whose mean is very close to the overall mean are left out, because for them the proportion is unstable.)

```{r echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE}
## proportion by which each class estimate is moved towards the overall mean
overall_mean <- fixef(lmm_model)[1]
shrinkage_data <- new_data |>
  filter(abs(predicted_lm - overall_mean) > 2) |>
  mutate(shrinkage = (predicted_lm - predicted_lmm) / (predicted_lm - overall_mean))
ggplot(shrinkage_data, aes(x = n_students, y = shrinkage)) +
  geom_point(alpha = 0.3) +
  labs(x = "Number of students in the class",
       y = "Shrinkage towards the overall mean (proportion)") +
  theme_minimal()
```

Classes with only a few students are pulled in a lot (more than half the way to the overall mean for classes of two); classes with many students are pulled in much less. With few students, a class mean is a noisy estimate, so the model borrows more information from the other classes.

```{r quiz-lmm-density-interpretation, echo=FALSE}
# LO: LO11.1.7
learnr::question(
  "After looking at the density distributions of the class-level predictions from the linear model (LM) and the linear mixed-effects model (LMM), what is the most appropriate interpretation?",
  
  learnr::answer(
    "The LMM predictions are more extreme than the LM predictions, indicating greater between-class variability."
  ),
  learnr::answer(
    "The LM predictions are tightly clustered around the overall mean, while the LMM predictions are widely spread."
  ),
  learnr::answer(
    "The LMM predictions show a narrower distribution than the LM predictions, reflecting shrinkage toward the overall mean.",
    correct = TRUE
  ),
  learnr::answer(
    "Both models produce identical distributions of predicted values."
  ),
  
  correct = "Correct! Partial pooling in the LMM pulls extreme class-level predictions toward the overall mean, resulting in a narrower distribution.",
  incorrect = "Focus on how partial pooling affects the spread of predictions across classes.",
  allow_retry = TRUE
)
```

```{r quiz-lmm-why-this-matters, echo=FALSE}
# LO: LO11.1.7
learnr::question(
  "Why does the difference between the LM and LMM predictions matter in practice?",
  
  learnr::answer(
    "Because mixed-effects models always give higher R-squared values than linear models."
  ),
  learnr::answer(
    "Because the LMM predictions are guaranteed to be unbiased for every individual class."
  ),  
  learnr::answer(
    "Because the LMM eliminates all between-class variability."
  ),
 
  learnr::answer(
    "Because partial pooling reduces the influence of random noise in small groups, leading to more reliable estimates and predictions.",
    correct = TRUE
  ),
 
  correct = "Correct! Partial pooling balances group-specific information with information from the whole dataset, reducing overfitting and improving reliability, especially for small groups.",
  incorrect = "Think about the role of noise, uncertainty, and sample size in estimating group-level effects.",
  allow_retry = TRUE
)
```

Think about how the same might apply in a random slopes model, where each group has its own slope. The same principles of partial pooling apply, with group-level slope estimates being shrunk toward the overall slope.

```{r quiz-lmm-random-slopes-contrast, echo=FALSE}
# LO: LO11.1.6, LO11.1.7
learnr::question(
  "Suppose we now fit a **random-slopes mixed-effects model**, allowing the relationship between an explanatory variable (e.g. study time) and the response (score) to vary among classes. How does this differ conceptually from estimating separate slopes for each class using a standard linear model (e.g. an ANCOVA-style interaction)?",
    learnr::answer(
    "The mixed-effects model partially pools the class-specific slopes toward a common average slope, whereas the linear model estimates each class slope independently.",
    correct = TRUE
  ),
 
  learnr::answer(
    "The mixed-effects model assumes that all classes have completely independent slopes, with no relationship among them."
  ),
  learnr::answer(
    "The mixed-effects model estimates a single common slope and ignores class-level variation."
  ),
 learnr::answer(
    "The mixed-effects model forces all class-specific slopes to be identical."
  ),
  
  correct = "Correct! In a random-slopes model, class-specific slopes are shrunk toward the overall slope, balancing within-class information with information from all classes.",
  incorrect = "Think about how partial pooling applies to slopes in the same way it applies to intercepts.",
  allow_retry = TRUE
)
```

Should class be a random effect or a fixed effect? It depends on your research question and the context of your study. If you are interested in making inferences about the specific classes in your dataset, treating class as a fixed effect may be appropriate. However, if you want to generalize your findings to a larger population of classes, treating class as a random effect is often more suitable.

Another way to think about this is how we are identifying (naming) the classes. If the names are arbitrary labels (e.g., class_1, class_2, etc.) that do not carry specific meaning, it is often more appropriate to treat class as a random effect. If instead of class names we had meaningful categories (e.g., different teaching methods), then treating that factor as a fixed effect would make more sense. Then it would be appropriate to make a mixed model with teaching method as a fixed effect and class (name) as a random effect.

## Practical part 2
<!-- LO: LO11.1.1, LO11.1.2, LO11.1.3, LO11.1.4, LO11.1.6, LO11.1.8, LO11.1.9, LO11.1.10, LO11.1.11, LO11.1.12, LO11.1.13 -->

**Sleep deprivation and reaction time: pseudoreplication, random effects and variance components.**

In a sleep-deprivation study, 18 people were allowed only 3 hours of sleep per night for 9 nights. Each day (day 0 to day 9) their average reaction time was measured. **Question: how much does reaction time increase with each day of sleep deprivation?**

The data are in the `lme4` package:

```{r echo=TRUE, eval = FALSE}
library(tidyverse)
library(lme4)
data(sleepstudy)
```

The variables are `Reaction` (average reaction time, ms), `Days` (days of sleep deprivation) and `Subject` (the person).

```{r LO-sleep-structure, echo=FALSE}
# LO: LO11.1.1, LO12.1.3
learnr::question(
  "There are 180 rows: 18 subjects measured on 10 days each. What is the structure of these data?",
  learnr::answer("Repeated measures: the 10 measurements of each subject are grouped, and measurements from the same person are likely to be more similar to each other than to measurements from other people.", correct = TRUE),
  learnr::answer("180 independent observations."),
  learnr::answer("18 independent observations, each of which is a single reaction time."),
  correct = "Correct! The independent units are the 18 subjects. The days are repeated measurements within subjects.",
  incorrect = "Think about who was measured, and how many times.",
  allow_retry = TRUE
)
```

**1. Look at the data.** Plot reaction time against days, with a separate panel for each subject:

```{r echo=TRUE, eval = FALSE}
ggplot(sleepstudy, aes(x = Days, y = Reaction)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ Subject)
```

Notice how subjects differ both in their reaction time at day 0 (their intercept) and in how quickly their reaction time increases (their slope).

**2. The wrong model: ignoring the subjects.**

```{r echo=TRUE, eval = FALSE}
m_sleep_lm <- lm(Reaction ~ Days, data = sleepstudy)
summary(m_sleep_lm)
```

```{r LO-sleep-lm-df, echo=FALSE}
# LO: LO11.1.4, LO11.1.2
learnr::question(
  "The linear model reports 178 residual degrees of freedom. Why is this a warning sign?",
  learnr::answer("Because there are only 18 independent units (subjects). The model treats the 180 repeated measurements as if they were independent, so it thinks it has much more information than it does (pseudoreplication). Its standard errors and p-values are not trustworthy.", correct = TRUE),
  learnr::answer("It is not a warning sign: 180 − 2 = 178 is exactly right."),
  learnr::answer("Because the degrees of freedom should be 180."),
  correct = "Correct! Whenever the residual degrees of freedom are much larger than the number of independent units in the study, suspect pseudoreplication.",
  incorrect = "Compare the residual degrees of freedom with the number of independent units in the study.",
  allow_retry = TRUE
)
```

**3. One honest solution: analyse one number per subject.** Fit a separate regression for each subject, keep the 18 slopes, and test whether their mean differs from zero:

```{r echo=TRUE, eval = FALSE}
subject_slopes <- sleepstudy |>
  group_by(Subject) |>
  summarise(slope = coef(lm(Reaction ~ Days))[2])
t.test(subject_slopes$slope)
```

This is valid (one value per independent unit), but it throws away information: it ignores how precisely each slope is estimated, and it cannot easily include other explanatory variables. Mixed models use all the data while respecting the grouping. Keep the result of this t-test (t ≈ 6.77 on 17 degrees of freedom) to compare with below.

**4. A random intercept model.**

```{r echo=TRUE, eval = FALSE}
library(lmerTest)  # adds degrees of freedom and p-values (Satterthwaite) to lmer output
m_sleep_ri <- lmer(Reaction ~ Days + (1 | Subject), data = sleepstudy)
summary(m_sleep_ri)
```

Look at the **Random effects** part of the output. It gives two variance components: the variance among subjects in their average reaction time (`Subject (Intercept)`), and the residual variance within subjects.

```{r LO-sleep-ri-sd-subject, echo=FALSE}
# LO: LO11.1.9
learnr::question_numeric(
  "What is the estimated standard deviation among subjects (the random intercept), in ms?\n\n(Round to one decimal place.)",
  learnr::answer(37.1, correct = TRUE),
  correct = "Correct! About 37.1 ms among subjects, compared with a residual standard deviation of about 31.0 ms within subjects.",
  incorrect = "Look at the 'Std.Dev.' column for Subject (Intercept) in the Random effects part of summary().",
  allow_retry = TRUE,
  tolerance = 0.1
)
```

```{r LO-sleep-icc, echo=FALSE}
# LO: LO11.1.10
learnr::question_numeric(
  "What proportion of the total variance (after accounting for Days) is due to differences among subjects? Calculate it from the two **variances** (not standard deviations): among-subject variance / (among-subject variance + residual variance).\n\n(Round to two decimal places.)",
  learnr::answer(0.59, correct = TRUE),
  correct = "Correct! 1378 / (1378 + 960) ≈ 0.59. About 59% of the variation in reaction time that is not explained by days of sleep deprivation is due to consistent differences among people. Biologically: people differ a lot in their baseline reaction time, so measurements from the same person are strongly correlated, which is why they must not be treated as independent.",
  incorrect = "Use the 'Variance' column, not the 'Std.Dev.' column.",
  allow_retry = TRUE,
  tolerance = 0.01
)
```

**5. A random intercept and random slope model.** The plot showed that subjects also differ in their slopes. Allow for this:

```{r echo=TRUE, eval = FALSE}
m_sleep_rs <- lmer(Reaction ~ Days + (Days | Subject), data = sleepstudy)
summary(m_sleep_rs)
```

```{r LO-sleep-rs-formula, echo=FALSE}
# LO: LO11.1.6
learnr::question(
  "Which statement about the formula `Reaction ~ Days + (Days | Subject)` is correct?",
  learnr::answer("It estimates an average (fixed) slope of Days, and lets both the intercept and the slope vary among subjects (random intercepts and random slopes).", correct = TRUE),
  learnr::answer("It estimates a separate fixed slope for each subject, like an interaction in lm()."),
  learnr::answer("It only lets the intercept vary among subjects."),
  correct = "Correct! (1 | Subject) would give random intercepts only. If subjects were nested in, say, laboratories, we would write (1 | Laboratory/Subject); if each subject were measured in several sessions that are shared by all subjects (crossed random effects), we would write (1 | Subject) + (1 | Session).",
  incorrect = "Look at what is to the left of the | inside the brackets.",
  allow_retry = TRUE
)
```

**6. Compare the fixed effect of Days across the models.** Fill in this table (on paper or in your script) for the three models: estimate of the Days slope, its standard error, and the degrees of freedom used for its test.

```{r LO-sleep-se-compare, echo=FALSE}
# LO: LO11.1.4, LO11.1.11, LO11.1.3
learnr::question(
  "The estimated slope is 10.47 ms per day in all three models. Its standard error is 1.24 (lm, 178 df), 0.80 (random intercepts, 161 df) and 1.55 (random intercepts and slopes, about 17 df). Which statements are correct?",
  learnr::answer("The random slope model gives the largest standard error, because it recognises that subjects differ in their slopes: the average slope is estimated from 18 subjects, so it has about 17 degrees of freedom.", correct = TRUE),
  learnr::answer("The random slope model's test (t ≈ 6.77, 17 df) matches the t-test on the 18 subject slopes from step 3: here the mixed model reaches the same conclusion while using all the data.", correct = TRUE),
  learnr::answer("Degrees of freedom for fixed effects in mixed models have to be approximated (here with Satterthwaite's method, from lmerTest), and the approximation reflects how many independent units carry information about each effect.", correct = TRUE),
  learnr::answer("The linear model is best because it has the most degrees of freedom."),
  correct = "Correct! Ignoring the grouping (lm) or ignoring variation in slopes (random intercepts only) makes us overconfident about the effect of sleep deprivation. The model that matches the structure of the data gives an honest standard error. (With balanced data like these, it agrees exactly with the analysis of one slope per subject.)",
  incorrect = "Compare the standard errors and degrees of freedom with the number of subjects, and with your t-test from step 3.",
  allow_retry = TRUE
)
```

```{r LO-sleep-random-slope-sd, echo=FALSE}
# LO: LO11.1.9, LO11.1.10
learnr::question_numeric(
  "In the random slope model, what is the standard deviation among subjects in their slope (ms per day)?\n\n(Round to one decimal place.)",
  learnr::answer(5.9, correct = TRUE),
  correct = "Correct! About 5.9 ms per day. So, while reaction time increases by about 10.5 ms per day on average, people differ: roughly 95% of people would have slopes between about 10.5 − 2 × 5.9 ≈ −1 and 10.5 + 2 × 5.9 ≈ 22 ms per day (if slopes are normally distributed). Some people are hardly affected; others are affected about twice as much as average.",
  incorrect = "Look at the 'Std.Dev.' of Days in the Random effects part of summary(m_sleep_rs).",
  allow_retry = TRUE,
  tolerance = 0.1
)
```

**7. Check the model.** Plot the residuals against the fitted values (`plot(m_sleep_rs)`) and make a QQ-plot of the residuals (`qqnorm(resid(m_sleep_rs))`).

```{r LO-sleep-limitations, echo=FALSE}
# LO: LO11.1.12
learnr::question(
  "Which are assumptions or limitations of this mixed model that you should keep in mind?",
  learnr::answer("It assumes the subject intercepts and slopes are drawn from normal distributions, and the residuals are normal with constant variance.", correct = TRUE),
  learnr::answer("It assumes the effect of days is linear over the 10 days.", correct = TRUE),
  learnr::answer("With only 18 subjects, the among-subject variances (especially of the slopes) are estimated imprecisely.", correct = TRUE),
  learnr::answer("It assumes that the 180 measurements are independent."),
  correct = "Correct! The mixed model removes the assumption that all measurements are independent (it models the dependence within subjects), but it brings its own assumptions, and variance components need a reasonable number of groups to be estimated well.",
  incorrect = "Think about which assumptions remain from linear models, what new ones the random effects bring, and how many groups there are.",
  allow_retry = TRUE
)
```

**8. Report.**

```{r LO-sleep-report, echo=FALSE}
# LO: LO11.1.13
learnr::question(
  "Which is the best reporting sentence?",
  learnr::answer("Reaction time increased significantly with days of sleep deprivation (p < 0.001, 178 df)."),
  learnr::answer("Reaction time increased by on average 10.5 ms per day of sleep deprivation (95% CI 7.4–13.5 ms; linear mixed model with random intercepts and slopes for the 18 subjects, t = 6.8, df = 17, p < 0.0001), although subjects varied considerably in their sensitivity (SD of slopes among subjects 5.9 ms per day).", correct = TRUE),
  learnr::answer("The random effects model was significant, with a variance of 1378."),
  correct = "Correct! It gives the effect size with its uncertainty, the model and its structure, a test with degrees of freedom that reflect the number of independent units, and what the variance components say biologically.",
  incorrect = "Look for the sentence that reports the effect with its uncertainty, uses honest degrees of freedom, and interprets the variation among subjects.",
  allow_retry = TRUE
)
```

(The confidence interval comes from `confint(m_sleep_rs, parm = "beta_", method = "Wald")`.)


## Practical part 3
<!-- LO: LO11.2.1, LO11.2.3, LO11.2.4, LO11.2.5, LO11.2.6 -->

The second part of the lecture was about what types of statistical methods might be used for questions and datasets we did not encounter during the course. Read the relevant sections of the course book (Chapter 11.2), then try these questions. There is no data analysis in this part.

```{r LO-whatnext-within-bio144, echo=FALSE}
# LO: LO11.2.1, LO12.1.2
learnr::question(
  "Which of these studies could be analysed with methods covered in BIO144?",
  learnr::answer("The number of seedlings in 60 plots in relation to light level (a Poisson or quasi-Poisson GLM).", correct = TRUE),
  learnr::answer("Whether each of 200 lizards was infected with a parasite, in relation to body size (a binomial GLM).", correct = TRUE),
  learnr::answer("Growth of 40 trees measured each year for 5 years, in relation to fertiliser treatment (a mixed model with tree as a random effect).", correct = TRUE),
  learnr::answer("The abundances of 30 species at 50 sites, to describe how community composition varies among sites (ordination).", correct = TRUE),
  learnr::answer("Daily air temperature measured at one weather station for 30 years, to forecast next year (time series analysis)."),
  correct = "Correct! The first four are covered by linear models, GLMs, mixed models and ordination. Long time series, where each observation depends on the previous ones, need time series methods.",
  incorrect = "Match each study to the type of response and data structure. Which one has strong dependence among successive observations over time?",
  allow_retry = TRUE
)
```

```{r LO-whatnext-beyond, echo=FALSE}
# LO: LO11.2.3, LO11.2.4
learnr::question(
  "A researcher records how many days each of 120 tagged fish survive after release, until the study ends after 200 days (some fish are still alive at the end). Which feature of these data calls for methods beyond BIO144, and which family of methods?",
  learnr::answer("A time-to-event response, where some survival times are only known to be longer than 200 days: survival analysis.", correct = TRUE),
  learnr::answer("A count response: a Poisson GLM."),
  learnr::answer("Many response variables: ordination."),
  correct = "Correct! Survival analysis handles time-to-event data, including individuals whose event had not happened by the end of the study (censored observations).",
  incorrect = "What kind of response is 'number of days until death', and what is special about the fish still alive at the end?",
  allow_retry = TRUE
)
```

```{r LO-whatnext-planning, echo=FALSE}
# LO: LO11.2.5, LO11.2.6
learnr::question(
  "Which statements about choosing and planning an analysis are correct?",
  learnr::answer("The analysis, including how many replicates are needed (power analysis), should be planned when the study is designed, before the data are collected.", correct = TRUE),
  learnr::answer("The research question should determine the analysis, not the techniques we happen to know.", correct = TRUE),
  learnr::answer("It is best to collect as much data as possible first and decide on the question and analysis afterwards."),
  learnr::answer("If we only know linear models, we should change the question so that a linear model can answer it."),
  correct = "Correct! Planning the analysis with the design avoids collecting data that cannot answer the question (or too few data to answer it).",
  incorrect = "Think about the first steps of the data analysis workflow from Chapter 1.",
  allow_retry = TRUE
)
```

## Weekly Quiz

```{r quiz-mm-why, echo=FALSE}
# LO: LO11.1.2
learnr::question(
  "What is the main reason for using a linear mixed-effects model instead of a standard linear model?",
  
  learnr::answer("To increase the number of explanatory variables."),
  learnr::answer(
    "To account for dependence among observations caused by grouping or repeated measurements.",
    correct = TRUE
  ),
  learnr::answer("To guarantee smaller p-values."),
  learnr::answer("To remove all between-group variation."),
  
  correct = "Correct! Mixed models explicitly account for non-independence due to grouping or repeated measures.",
  incorrect = "Think about the assumptions of independence in linear models.",
  allow_retry = TRUE
)
```


```{r quiz-mm-partial-pooling, echo=FALSE}
# LO: LO11.1.7
learnr::question(
  "What does *partial pooling* mean in the context of mixed-effects models?",
  
  learnr::answer("All groups are forced to have exactly the same mean."),
  learnr::answer("Each group is analysed completely independently."),
  learnr::answer(
    "Group-level estimates are shrunk toward the overall mean, depending on data availability and variability.",
    correct = TRUE
  ),
  learnr::answer("Only the largest groups influence the model."),
  
  correct = "Correct! Partial pooling balances group-specific information with information from the whole dataset.",
  incorrect = "Think about how information is shared across groups.",
  allow_retry = TRUE
)
```


```{r quiz-mm-shrinkage, echo=FALSE}
# LO: LO11.1.7
learnr::question(
  "Which groups experience the strongest shrinkage in a mixed-effects model?",
  
  learnr::answer(
    "Groups with few observations, all else being equal.",
    correct = TRUE
  ),
  learnr::answer("Groups with many observations."),
  learnr::answer("Only groups with extreme means, regardless of sample size."),
  learnr::answer("All groups equally."),
  
  correct = "Correct! Groups with little data provide weaker evidence and are therefore shrunk more toward the overall mean.",
  incorrect = "Think about how sample size affects certainty.",
  allow_retry = TRUE
)
```



```{r quiz-mm-random-intercept, echo=FALSE}
# LO: LO11.1.6
learnr::question(
  "Which model formula specifies a **random intercept for class** in `lme4`?",
  
  learnr::answer("score ~ class"),
  learnr::answer(
    "score ~ 1 + (1 | class)",
    correct = TRUE
  ),
  learnr::answer("score ~ (class | 1)"),
  learnr::answer("score ~ class + error(class)"),
  
  correct = "Correct! (1 | class) specifies a random intercept for class.",
  incorrect = "Look for the syntax that places class inside parentheses after |.",
  allow_retry = TRUE
)
```


```{r quiz-mm-random-slopes, echo=FALSE}
# LO: LO11.1.6
learnr::question(
  "How does a random-slopes model differ from fitting separate slopes using an interaction in a linear model?",
  
  learnr::answer("Random slopes force all groups to have identical slopes."),
  learnr::answer("Random slopes ignore group-level variation."),
  learnr::answer("There is no difference; they are mathematically identical."),
   learnr::answer(
    "Random slopes partially pool group-specific slopes toward a common average slope.",
    correct = TRUE
  ),
 
  correct = "Correct! Random slopes allow slopes to vary by group while sharing information across groups.",
  incorrect = "Think about partial pooling applied to slopes.",
  allow_retry = TRUE
)
```


```{r quiz-mm-df, echo=FALSE}
# LO: LO11.1.11
learnr::question(
  "Why are degrees of freedom in mixed-effects models less straightforward than in standard linear models?",
  
  learnr::answer("Because mixed models always have infinite degrees of freedom."),
  learnr::answer(
    "Because variance components and partial pooling complicate how uncertainty is allocated.",
    correct = TRUE
  ),
  learnr::answer("Because R does not compute them correctly."),
  learnr::answer("Because mixed models do not estimate parameters."),
  
  correct = "Correct! Random effects and variance components make the effective degrees of freedom more complex.",
  incorrect = "Think about how uncertainty is shared across levels of the model.",
  allow_retry = TRUE
)
```

```{r quiz-mm-fixed-effects, echo=FALSE}
# LO: LO11.1.5
learnr::question(
  "Which of the following are **fixed effects**?",
  
  learnr::answer("Patient", correct = FALSE),
  learnr::answer("Tree", correct = FALSE),
  learnr::answer("Amount of rain", correct = TRUE),
  learnr::answer("Distance", correct = TRUE),
  learnr::answer("Colour", correct = TRUE),
  
  correct = "Correct. Fixed effects are variables whose specific values are of interest and repeatable.",
  incorrect = "Remember: fixed effects are repeatable quantities or categories of interest, not individual units.",
  allow_retry = TRUE
)
```


```{r quiz-mm-random-effects, echo=FALSE}
# LO: LO11.1.5
learnr::question(
  "Which of these are **random effects**?",
  
  learnr::answer("Hospital", correct = TRUE),
  learnr::answer("Nest", correct = TRUE),
  learnr::answer("Laboratory", correct = TRUE),
  learnr::answer("Observer", correct = TRUE),
  learnr::answer("Drug concentration", correct = FALSE),
  
  correct = "Correct. These represent sources of random variation sampled from a larger population.",
  incorrect = "Random effects represent sampled levels, not controlled treatment values.",
  allow_retry = TRUE
)
```


```{r quiz-mm-why-multilevel, echo=FALSE}
# LO: LO11.1.3
learnr::question(
  "Why should we prefer a **multilevel (mixed-effects) model** rather than simply averaging across groups? (Multiple answers may be correct.)",
  
  learnr::answer("To avoid false confidence about our data.", correct = TRUE),
  learnr::answer("Because multilevel models are easier to interpret and communicate.", correct = FALSE),
  learnr::answer("Averaging requires an arbitrary decision about which type of average to use.", correct = TRUE),
  learnr::answer("Because we retain variation that may indeed be of some interest.", correct = TRUE),
  
  correct = "Correct. Multilevel models retain meaningful variation and avoid overconfidence caused by averaging.",
  incorrect = "Think about what information is lost when averaging across groups.",
  allow_retry = TRUE
)
```





```{r quiz-what-next-nonlinear, echo=FALSE}
# LO: LO11.2.4
learnr::question(
  "If you suspect a smooth but non-linear relationship between an explanatory variable and a response, which approach introduced in this section would often be appropriate?",
  
  learnr::answer("Breakpoint analysis."),
  learnr::answer("Non-parametric rank tests."),
  learnr::answer(
    "Generalized Additive Models (GAMs).",
    correct = TRUE
  ),
  learnr::answer("Meta-analysis."),
  
  correct = "Correct! GAMs allow flexible, smooth non-linear relationships while remaining interpretable.",
  incorrect = "Focus on methods designed specifically for smooth non-linear effects.",
  allow_retry = TRUE
)
```

```{r quiz-time-series-biological, echo=FALSE}
# LO: LO11.2.2, LO11.2.4
learnr::question(
  "A biologist measures the population size of a plankton species in a lake every week for five years, with the goal of understanding long-term trends and seasonal dynamics. Which type of analysis is most appropriate for these data?",
  
  learnr::answer("A standard linear regression, because population size is a continuous variable."),
  learnr::answer(
    "A time series analysis, because measurements are ordered in time and successive observations are likely to be dependent.",
    correct = TRUE
  ),
  learnr::answer("A two-sample t-test, comparing early and late years."),
  learnr::answer("A one-way ANOVA with week as a factor."),
  
  correct = "Correct! Repeated measurements over time are typically not independent, making time series analysis appropriate.",
  incorrect = "Focus on the fact that the same system is measured repeatedly over time.",
  allow_retry = TRUE
)
```

```{r quiz-sem-biological, echo=FALSE}
# LO: LO11.2.4
learnr::question(
  "An ecologist is studying how nutrient availability affects plant growth. Nutrients influence leaf nitrogen content, which affects photosynthetic rate, which in turn affects total biomass. At the same time, nutrient availability may also have a direct effect on biomass. The ecologist wants to quantify both direct and indirect effects in a single analysis. Which type of analysis is most appropriate?",
  
  learnr::answer("Multiple linear regression, with biomass as the response."),
  learnr::answer("A one-way ANOVA comparing nutrient treatments."),
  learnr::answer(
    "Structural Equation Modelling (SEM), because it allows modelling of multiple interconnected relationships and indirect effects.",
    correct = TRUE
  ),
  learnr::answer("Time series analysis of biomass."),
  
  correct = "Correct! SEM is designed for analysing systems of variables with both direct and indirect relationships, and where variables can be both predictors and responses.",
  incorrect = "Think about methods that can represent causal pathways and indirect effects explicitly.",
  allow_retry = TRUE
)
```
